Prenatal conditions do not affect brain physiology and learning in a lizard

### 2025-03-23 ms.qmd
updates

---
title: "Spatial learning and brain mitochondria are robust to prenatal conditions in the delicate skink"
docx: 
bibliography: "./bib/refs.bib"
csl: "./bib/biology-letters.csl"
reference-doc: "./bib/tmpl.docx"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

## Abstract
Learning is highly sensitive to environmental conditions, including early thermal environments and prenatal exposure to glucocorticoids. These factors influence mitochondrial activity, which plays a crucial role in neural development and synaptic plasticity, ultimately shaping cognitive abilities. Here, we investigated the interactive effects of prenatal temperature and corticosterone (CORT) — the primary glucocorticoid in reptiles — on spatial learning and brain mitochondrial activity in the delicate skink (_Lampropholis delicata_). We supplemented eggs with either CORT or a control vehicle and incubating them under two thermal regimes: cold (23 ± 3°C) or hot (28 ± 3°C). Juveniles were then tested in a spatial learning task over 40 days, after which mitochondrial activity in the medial cortex was assessed. Our results revealed that both mitochondrial physiology and spatial learning in _L. delicata_ are robust to prenatal temperature and CORT exposure. In addition, we found no significant relationship between mitochondrial function and cognitive abilities. However, mitochondrial potential increased ROS levels, which could impact oxidative stress under high metabolic demands or low antioxidant capacity. These findings highlight the cognitive and physiological resilience of _L. delicata_ to early-life environmental challenges. Future research should explore whether this robustness extends to other brain regions and life stages.

@@

Learning - the acquisition and consolidation of new information [@dukas_evolutionary_2004] - enables individuals to create new associations between events, which can be essential for coping with environmental changes [@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013]. However, the capacity to form these associations varies among individuals, potentially affecting their responses to environmental challenges [@ward2016ecological; @welklin2024spatial]. Animals that learn faster may exploit novel resources better or may start responding to new threats more efficiently than slow learners [@ward2016ecological]. Consequently, individual differences in learning abilities can have profound effects on survival and reproductive output, ultimately influencing population growth rates and stability [@ward2016ecological; @welklin2024spatial]. Understanding the mechanisms that drive these differences is therefore crucial for predicting how populations will respond to novel conditions.

Factors like age, sex, or early- life conditions can have important effects on learning abilities [@szuran_water_1994; @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @carazo_sex_2014; @noble_age-dependent_2014]. Early developmental environment is particularly influential, as the brain is highly sensitive to environmental inputs during this period [@zhu_prenatal_2004]. In fact, early experiences can alter neurotransmitter production [@amani2021perinatal], gene expression [@zhou2020effects], or brain structure [@amiel_effects_2017], with lasting effects on cognition. Among the physiological mechanisms underlying learning, mitochondrial activity is considered a key factor [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. The neural processes involved learning impose substantial energetic demands [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], making efficient mitochondrial respiration essential [@du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Additionally, learning depends on a dense network of functional neurons [@lefebvre_taxonomic_2011; @amiel_effects_2017], which can be compromised by excessive oxidative stress byproduct of mitochondrial function [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. Studies on mammals have shown the pervassive effects of mitochondrial physiology on cognitive abilities [@zhu_prenatal_2004; @hara_presynaptic_2014; @cao2014ampk]. However, the extent to which these effects are generalizable to other taxa remains largely unexplored.

Since mitochondria are maternally inherited, maternal condition plays a fundamental role in shaping offspring mitochondrial activity [@picard_mitochondria_2014]. In particular, maternal stress can strongly influence how mitochondria operate in the offspring [@zhu_prenatal_2004]. Stressful situations elevate glucocorticoids (GCs) [@sapolsky_how_2000], which can be transmitted to the developing embryo [@uller_sex-specific_2009] altering mitochondrial physiology through transgenerational effects [@picard_mitochondria_2014]. For instance, maternal stress has been shown to contribute significantly to oxidative stress in the brain of rats (_Rattus norvegicus_) [@zhu_prenatal_2004; @haussmann_embryonic_2012; @cao2014ampk].

Temperature can be a significant source of maternal stress in ectotherms. Thermal environments outside of the optimal range can elevate GCs in the mother, which can be passed to the offspring [see @Crino_2023]. In addition, temperature can also directly influence offspring development, particularly during early life [@noble_developmental_2018; @crino2024eggs]. The prenatal thermal environment plays a crucial role in shaping mitochondrial function, affecting energy metabolism and oxidative stress [@stier2022experimental; @crino2024eggs]. For instance, hatchlings of the red-eared slider turtle (_Trachemys scripta elegans_) incubated at 26.5 ºC exhibited greater oxidative damage than those incubated at temperatures 0.7 to 1.2 ºC warmer [@treidel2016temperature]. Thus, the combined effects of prenatal GCs and temperature may profoundly influence mitochondrial function, with important consequences for brain development and cognition. However, the extent to which prenatal GCs and temperature interact to shape cognitive abilities via mitochondrial physiology remains largely unknown.

Here, we examined how prenatal temperature and corticosterone (CORT) — the primary GC in reptiles — affect brain mitochondrial physiology and spatial learning in the delicate skink (_Lampropholis delicata_). We hypothesized that prenatal CORT and temperature would influence learning abilities through impacts on brain mitochondrial activity. Specifically, we predicted that prenatal CORT would decrease energy production while increasing oxidative damage [@costantini2011meta; @gong_chronic_2011; but see @crino2024eggs]. Similarly, high temperatures would decrease mitochondrial efficienty [@zavorka_climate_2021; @crino2024eggs], but with the potential to reduce oxidative stress [@treidel2016temperature]. We further predicted that the combined effects of CORT and temperature would lead to complex interactions, with both factors negatively impacting mitochondrial efficiency but having opposite effects on oxidative stress. These effects would lead to differences in learning abilities, that could be affected by the balance between energy production and oxidative stress [@du_dynamic_2009; @picard_energetic_2018; @alexandrov_neuronal_2022]. By examining these interactions, we aim to clarify how prenatal environmental factors shape learning abilities through mitochondrial function, shedding light on the role of early-life conditions in cognitive development and its mechanismsm.

@@

#### Animals' origin and collection
Subjects came from a breeding colony established in the laboratory since 2019. It consisted in 270 adults living in groups of two males and four females. Eggs were collected from these groups between mid-Novemnber 2023 to mid-January 2024. After collecting the eggs, we treated them with CORT or vehicle and incubated them under two different temperature regimes (see below). Clutch and egg identity were asigned immediately after collection, and the eggs were incubated in individual cups until hatching. Hatchlings were also kept in individual enclosures until the end of the experiment. For details on husbandry and breeding conditions, see Methods 1: Animal husbandry in Supplementary Material.

#### CORT and Temperature manipulation
We tested the combined effects of prenatal CORT and temperature by elevating CORT concentrations in eggs and then incubating them under one of two temperature regimes (Cold - 23 ± 3 ºC or Hot - 28 ± 3 ºC) in a 2x2 factorial design. We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. CORT-treated eggs were topically supplied with 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), while Control eggs received an equal volume of 100% Ethanol. CORT dose increased mean yolk CORT levels ~3.7x higher than control eggs in previous studies [@crino2024eggs]. Eggs were then incubated in one of the two previously mentioned temperature regimes that are within the natural limits in _L. delicata_ [@cheetham2011embryonic].

#### Learning  
The spatial learning task involved training lizards to navigate a 6-arm maze to reach an exit connected to a transport box that allowed us to return the lizards to their enclosure without further contact. In each trial, the lizards were placed by hand in the center of maze and left to acclimate for two minutes. During acclimatation, the central area was surrounded by a yellow device mounted on a pulley system. At the start of each trial, this device was lifted to startle the lizard. If the lizard did not immediately choose an arm, it was gently prodded with a brush at the end of the tail. Once the lizard made a choice, the brush was used to encourage movement but without guiding it towards any specific arm. If the lizard did not choose correctly after 20 errors — at which point it typically ceased responding — the lizard was gently guided to the correct arm.

We assessed lizards' spatial learning using external cues for navigation while avoiding intra-maze cues. To avoid subtle intra-maze cues, the maze was replaced every three trials with one of four identical mazes. Each replacement preserved the correct arm’s orientation and the maze’s position within the room, ensuring consistency for each individual. Additionally, the maze was cleaned with 70% ethanol between trials to avoid the influence of chemical cues. To control for potential side biases, the correct arm was randomly assigned to one of the six arms for each trial. We employed four different maze orientations (North-West, South-West, South, and South-East), and the number of lizards assigned to each orientation was counterbalanced across treatments.

@@

#### Brain mitochondrial activity
Immediately following the completion of the tests, we asessed mitochondrial physiology in the brain of all remaining lizards (see below). We euthanized lizards via peritoneal injection 10 mg/kg of a 10 mg/mL alfaxan solution (a potent anesthetic), followed by decapitation. Before decapitation we evaluated lizard's righting response and the pinching reflex in one of the front limbs.

After decapitation, we extracted the medial cortext in the telencephalon. We selected the medial cortex because is considered homologous to the mammalian hippocampus, where spatial cognition is encoded [@rodríguez2002conservation; @naumann2015reptilian]. The tissue was transferred to 1xPBS solution and then homogenized mechanically using a 100 µm mesh filter (pluriStrainer). The resultant homogenate was divided into two aliquotes: one was used fresh for mitochondrial density, membrane potential, and ROS assays; and the other was cryopreserved for later measurements of DNA damage and lipid peroxidation.

Fresh homogenate suspensions were stained with 5 µM MitoTracker Deep Red FM, 2.5 µM MitoTracker Orange CMTMRos, and 50 µM MitoSOX Red. We used these fluorescent probes to assess mitochondrial density, mitochondrial membrane potential, and superoxide (ROS) production, respectively. We also added 5 µL of #µM Hoechst 33342 Nuclear Viability Dye to each sample to differentiate viable cells from debris. These samples were analyzed in a flow cytometer within 2 hours of brain extraction.

The aliquotes reserved for oxidative damage assays were stained with 20 µL of #µM Hoechst 33342 Nuclear Viability Dye and 20 µL of 100 µM BODIPY 665/676 Lipid Peroxidation Sensor before cryopreservation. These dyes were used to measure cell viability and lipid peroxidation, respectively. The samples were then fixed in 1% Neutral-Buffered Formalin, washed, and preserved in a 1X Tris-EDTA solution with 10% DMSO at -20 °C. On the day of the oxidative damage assays, the samples were thawed, and DMSO was removed. Then, cell membranes were permeabilized in 200 µL of 1X PBS containing 5mM (6.14 mg/mL) digitonin. Following membrane permeabilization we stained the samples with 20µL of 70µM 8-OHdG Polyclonal Antibody - a marker of DNA damage - and allowed them to incubate overnight (~12 hours). The following day we counterstained the cells with 20µL of #µM (100µg/mL) H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 and analyzed the samples in the the flow cytometer. Oxidative damage assays were performed within 6 months of the initial analysis of fresh samples.

Flow cytometry was performed on a Becton Dickson LSRFortessa X-20 flow cytometer using the default wavelength filters. The detectors and voltage settings for each assay were determined during pilot trials and and remained consistent throughout the experiment. Data was processed using FlowJo (v. 10.1) software. We obtained the mean fluorescent intensity for mitochondrial density, mitochondrial potential, ROS production, DNA damage, and lipid peroxidation. For further details on the homogenization, staining, or flow cytometry assays, see Methods 2: flow cytometry in Supplementary Material.

#### Statistical analyses
We performed all the analyses using brm models from stan [@stan] on R version 4.4.0 [@R]. We run a series of univariate and multivariate models to test the effects of early envirnoment on each variable separately plus the relationships between physiology and learning. All models consisted of four MCMC chains of 8000 iterations, with a warm up interval of 2000 iterations.

Univariate models were used to test the effects of early environment on each of the variables recorded: mitochondrial density, mitochondrial potential, ROS, DNA damage, lipid peroxidation, and the number of errors as a measure of learning. We included in all the models the hormone treatment (CORT vs Control), incubation temperature (Cold vs Hot) and their interaction. Sex and age were included in preliminary models and excluded from the final models when they were not significant. For all univariate models, clutch identity was included as a random factor. Mitochondrial related variables (density, potential, ROS, DNA damage, and lipid peroxidation) where log-transformed when necessary and all were mean-centered and standardized by dividing them by two times the standard deviation [@gelman2008scaling]. These variables were considered to follow a normal distribution.

The number of errors was modeled as a function of trial, cort, temperature, and the triple interaction. We included in this model the trial within each level of lizard identity as a random slope. The error structure was modelled using a negative binomial distribution with a logit link function [negbinomial(link = "log")]. Otherwise, the procedure was as in the other models.

We used the posterior distributions of parameters from these models to test for differences between treatments. Learning slopes were obtained using the ‘trial’ estimates and its interaction with hormone and temperature treatments. Slope estimates <0 were considered as evidence of learning. pMCMC were used to test the hypothesis that slopes and slope contrasts are different from zero. We considered an effect statistically significant when pMCMC <0.05.

To explore links between mitochondrial function and detection latency, we fitted a multivariate model on brm. We included in the model all the variables of interest and their interactions structured following a specific hypothesis (see @fig-sem). Individual measurements for learning performance were obtained by modelling errors as a function of trial. We included the trial within each level of lizard identity as a random slope, which was used to obtain the learning slopes for each individual. These learning slopes, as well as the rest of variables, were standardized as before (see above). We assumed a gaussian family for all the variables. Factors found as non-significant in the univariate models were excluded. Clutch identity remained a random factor. Direct, indirect, and total effects were derived from posterior estimates following a Structural Equations Modelling approach [@kline2005principles] (see Table S4).

One of the plates used to analyze oxidative damage malfunctioned, compromising data quality. This data was excluded from the final analysis and, as a result, the sample size for DNA damage and lipid peroxidation is smaller than for the other variables. For the multivariate model, we imputed the missing values. The final sample sizes are presented below.

@@

We started with 80 lizards - 20 per treatment - coming from `r total_clutches` clutches. However, due to natural mortality (n = 2), only 79 lizards were included in the learning analyses, and 78 in the mitochondrial analyses. Besides, due to malfunctioning of the flow cytometer, we ended up with 40 inividuals (10 per treatment) for DNA damage and lipid peroxidation. Final sample sizes per treatment and species are disclosed on @fig-learning and @fig-results_mit.

@@

The multivariate model showed that mitochondrial physiology was not related to learning abilities (see @fig-sem and Table S4). We did not find significant relationships between ROS and oxidative damage or mitochondrial density and ROS (see @fig-sem and Table S4). However, we found that ROS production increased with mitochondrial potential (β = `r format_dec(mean(coeff_potential_ros), 3)`, 95% CI = [`r format_dec(quantile(coeff_potential_ros, 0.025), 3)`, `r format_dec(quantile(coeff_potential_ros, 0.975), 3)`], pMCMC `r format_p(pmcmc(coeff_potential_ros), 3, equal = TRUE)`).

>>>>

---
title: "Prenatal conditions do not affect brain physiology and learning in a lizard"
docx: 
bibliography: "./bib/refs.bib"
csl: "./bib/biology-letters.csl"
reference-doc: "./bib/tmpl.docx"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Figure    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Figure   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

## Abstract
Early environmental factors like heat or stress hormones can impair learning through brain metabolic function, crucial for neural development and synaptic plasticity. However, whether early environments always result in cognitive impairment through changes in neural physiology is not well-established outside of a few model systems. Here, we investigated the effects of prenatal temperature and corticosterone (CORT) on brain mitochondrial activity and spatial learning in the delicate skink (_Lampropholis delicata_). We treated eggs with either CORT or a control vehicle and incubated at cold (23 ± 3°C) or hot (28 ± 3°C) temperatures. Juveniles were tested in a spatial learning task over 40 days after which mitochondrial function in the medial cortex was assessed. Despite among-individual variation in learning ability, mitochondrial physiology and spatial learning in _L. delicata_ remained robust to prenatal temperature and CORT exposure. No significant relationship was found between mitochondrial function and cognitive performance, contrary to predictions. Additionally, increased metabolic capacity correlated with higher ROS production but did not affect oxidative damage, possibly due to protective mechanisms. These findings highlight the physiological and cognitive resilience of _L. delicata_ to early-life challenges. Future research should explore whether this robustness extends to other brain regions, cognitive domains, and life stages.

@@

Learning - the acquisition and consolidation of new information [@dukas_evolutionary_2004] - enables animals to create new associations between events, which can be essential for coping with environmental change [@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013]. However, the capacity to form these associations varies among individuals, potentially affecting their responses to environmental challenges [@ward2016ecological; @welklin2024spatial]. Learning faster may imply exploiting novel resources better or responding to new threats more efficiently [@ward2016ecological]. Consequently, individual differences in learning abilities can have profound effects on fitness, ultimately influencing population growth rates and stability [@ward2016ecological; @welklin2024spatial]. Understanding the mechanisms that drive these differences is therefore crucial for predicting how populations will respond to novel conditions.

Factors like age, sex, or early-life conditions can have important effects on learning abilities [@szuran_water_1994; @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @carazo_sex_2014; @noble_age-dependent_2014]. Adverse developmental environments are particularly influential, as the brain is highly sensitive to environmental inputs during early life [@zhu_prenatal_2004]. In fact, early experiences can alter neurotransmitter production [@amani2021perinatal], gene expression [@zhou2020effects], or brain structure [@amiel_effects_2017], with lasting effects on cognition. Among the physiological mechanisms underlying learning, mitochondrial activity is considered a key factor [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. The neural processes involved in learning impose substantial energetic demands [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], making efficient mitochondrial respiration essential [@du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Additionally, learning depends on a dense network of functional neurons [@lefebvre_taxonomic_2011; @amiel_effects_2017], which can be compromised by excessive oxidative stress - a byproduct of mitochondrial metabolism [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. Studies on mammals have shown the pervasive effects of mitochondrial physiology on cognitive abilities [@zhu_prenatal_2004; @hara_presynaptic_2014; @cao2014ampk]. However, the extent to which these effects are generalizable to other taxa remains largely unexplored.

Since mitochondria are maternally inherited, maternal condition plays a fundamental role in shaping offspring mitochondrial activity [@picard_mitochondria_2014]. Maternal stress can also influence how mitochondria operate in offspring [@zhu_prenatal_2004]. Stressful situations experienced by mothers can elevate glucocorticoids (GCs) [@sapolsky_how_2000], which can be transmitted to the developing embryos [@uller_sex-specific_2009] altering mitochondrial physiology through transgenerational effects [@picard_mitochondria_2014]. For instance, maternal stress has been shown to contribute significantly to oxidative stress in the brain of rats (_Rattus norvegicus_) with impacts on spatial learning abilities [@zhu_prenatal_2004; @haussmann_embryonic_2012; @cao2014ampk].

Temperature is also a significant source of maternal stress in ectotherms. Thermal environments outside of the optimal range can elevate GCs in mothers, which can be passed to the offspring [see @Crino_2023]. Additionally, temperature can also directly influence offspring development, particularly during early life [@noble_developmental_2018; @crino2024eggs]. The prenatal thermal environment plays a crucial role in shaping mitochondrial function, affecting energy metabolism and oxidative stress [@stier2022experimental; @crino2024eggs]. Thus, the combined effects of prenatal GCs and temperature may profoundly influence mitochondrial function, with important consequences for brain development and cognition. However, the extent to which prenatal GCs and temperature interact to shape cognitive abilities via mitochondrial physiology remains largely unknown outside of a few model species [@zhu_prenatal_2004; @haussmann_embryonic_2012; @cao2014ampk].

Here, we examined how prenatal temperature and corticosterone (CORT) — the primary GC in reptiles — affect brain mitochondrial physiology and spatial learning in the delicate skink (_Lampropholis delicata_). We hypothesized that prenatal CORT and temperature would influence learning abilities through impacts on brain mitochondrial activity. Specifically, we predicted that prenatal CORT would decrease energy production while increasing oxidative damage [@costantini2011meta; @gong_chronic_2011; but see @crino2024eggs], especially if CORT can alter cellular components increasing ROS production without enhancing ATP synthesis or making them more vulnerable to oxidative damage. Similarly, high temperatures would decrease mitochondrial efficiency [@zavorka_climate_2021; @crino2024eggs], but with the potential to reduce oxidative stress [@treidel2016temperature]. We further predicted that the combined effects of CORT and temperature would lead to complex interactions, with both factors negatively impacting mitochondrial efficiency but having opposite effects on oxidative stress. These effects would lead to differences in learning abilities, that could be affected by the balance between energy production and oxidative stress [@du_dynamic_2009; @picard_energetic_2018; @alexandrov_neuronal_2022]. By examining these interactions, we aim to clarify how prenatal environmental factors shape learning abilities through mitochondrial function, shedding light on the role of early-life conditions in cognitive development and associated mechanismsm.

@@

#### Experimental animals
Lizards came from a breeding colony established in the laboratory in 2019 from wild populations in Sydney, Australia. The colony consisted of 270 adults living in groups of two males and four females. Eggs were collected from these groups between November 2023 to January 2024. After collecting the eggs, we treated them with CORT or a vehicle control and incubated them under two different temperature regimes (see below). Clutch and egg identity were assigned immediately after egg collection, and the eggs were incubated in individual cups until hatching. Hatchlings were also kept in individual enclosures until the end of the experiment. For details on husbandry and breeding conditions see _Methods 1: Animal husbandry in Supplementary Material_.

#### Corticosterone and temperature manipulations
We tested the combined effects of prenatal CORT and temperature by elevating CORT concentrations in eggs and then incubating them under one of two temperature regimes (Cold - 23 ± 3 ºC or Hot - 28 ± 3 ºC). We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. CORT-treated eggs were topically supplied with 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final concentration of 10 pg CORT/mL (CORT treatment), while Control eggs received an equal volume of 100% Ethanol. CORT dose increased mean yolk CORT levels ~3.7x higher than control eggs in previous studies [@crino2024eggs]. Eggs were then incubated in one of the two previously mentioned temperature regimes that are within the natural limits of nest temperatures in _L. delicata_ [@cheetham2011embryonic].

#### Spatial learning  
The spatial learning task involved training lizards to navigate a 6-arm maze to reach an exit connected to a transport box that allowed us to return the lizards to their enclosure without further contact. In each trial, lizards were placed by hand in the center of maze and left to acclimatise for two minutes. During acclimatisation, the central area was surrounded by a yellow device mounted on a pulley system. At the start of each trial, this device was lifted to startle the lizard. If the lizard did not immediately choose an arm, it was gently prodded with a brush at the end of the tail. Once the lizard made a choice, the brush was used to encourage movement but without guiding it towards any specific arm. If the lizard did not choose correctly after 20 errors — at which point it typically ceased responding — the lizard was gently guided to the correct arm.

We assessed lizards' spatial learning using external cues for navigation while avoiding intra-maze cues. Subtle intra-maze cues were avoided by replacing the maze every three trials with one of four identical mazes. Each replacement preserved the correct arm’s orientation and the maze’s position within the room, ensuring consistency for each individual. Additionally, the maze was cleaned with 70% ethanol between trials to avoid the influence of chemical cues. To control for potential side biases, the correct arm was randomly assigned to one of the six arms for each trial. We employed four different maze orientations counterbalancing the number of lizards assigned to each orientation across treatments.

@@

#### Brain mitochondrial function
Immediately following the completion of the tests, we quantified mitochondrial physiology in the brain of lizards. We euthanized lizards via peritoneal injection 10 mg/kg of a 10 mg/mL alfaxan solution (a potent anesthetic), followed by decapitation. Before decapitation we evaluated lizard righting responses and the pinching reflexes in one of the front limbs.

We extracted the medial cortex in the telencephalon as this brain region is considered homologous to the mammalian hippocampus, where spatial cognition is encoded [@rodríguez2002conservation; @naumann2015reptilian]. The tissue was transferred to 1X PBS solution and then homogenized mechanically using a 100 µm mesh filter (pluriStrainer). The resultant homogenate was divided into two aliquots: one was used fresh for measuring mitochondrial density, membrane potential - a metric of mitochondrial metabolic capacity [@martinez2016tca] - and superoxide (ROS) production; and the other was cryopreserved for later measurements of DNA damage and lipid peroxidation.

Fresh homogenate suspensions were stained with 5 µM MitoTracker Deep Red FM, 2.5 µM MitoTracker Orange CMTMRos, and 50 µM MitoSOX Red. We used these fluorescent probes to assess mitochondrial density, membrane potential, and ROS, respectively. We also added 5 µL of 10 µg/mL Hoechst 33342 Nuclear Viability Dye to each sample to differentiate viable cells from debris. These samples were analyzed in a flow cytometer within 2 hours of brain extraction.

The aliquots reserved for oxidative damage assays were stained with 20 µL of 10 µg/mL Hoechst 33342 Nuclear Viability Dye and 20 µL of 100 µM BODIPY 665/676 Lipid Peroxidation Sensor before cryopreservation. These dyes were used to measure cell viability and lipid peroxidation, respectively. The samples were then fixed in 1% Neutral-Buffered Formalin, washed, and preserved in a 1X Tris-EDTA solution with 10% DMSO at -20 °C. On the day of the oxidative damage assays, the samples were thawed, and DMSO was removed. Then, cell membranes were permeabilized in 200 µL of 1X PBS containing 5mM (6.14 mg/mL) digitonin. Following permeabilization we stained the samples with 20µL of 70µM 8-OHdG Polyclonal Antibody - a marker of DNA damage - and allowed them to incubate overnight (~12 hours). The following day we counterstained the cells with 20µL of 100 µg/mL H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 and analyzed the samples using the flow cytometer. Oxidative damage assays were performed within 6 months of the initial analysis of fresh samples.

Flow cytometry was performed on a Becton Dickson LSRFortessa X-20 flow cytometer using the default wavelength filters. Detectors and voltage settings for each assay were determined during pilot trials and remained consistent throughout the experiment. Data was processed using FlowJo (v. 10.1) software. We obtained the mean fluorescent intensity for mitochondrial density, metabolic capacity, ROS, DNA damage, and lipid peroxidation. For further details on the homogenization, staining, or flow cytometry assays see _Methods 2: flow cytometry in Supplementary Material_. Sample sizes for DNA damage and lipid peroxidation were smaller than for the other variables because of a plater reader malfunction, however, data was missing randomly and we dealt with missing data in our analyses (see below).

#### Statistical analyses
We performed all analyses using the _brm_ package [@burkner2017brms] which fits Bayesian multilevel models with Stan [@stan] with R version 4.4.0 [@R]. We ran a series of univariate and multivariate models to test the effects of early envinonment on each variable separately plus quantify the relationships between physiology and learning. All models consisted of four MCMC chains of 8000 iterations, with a warmup interval of 2000 iterations.

Univariate models were used to test the effects of early environment on each of the variables recorded: mitochondrial density, metabolic capacity, ROS, DNA damage, lipid peroxidation, and the number of errors as a measure of learning. We included in all the models the hormone treatment (CORT vs Control), incubation temperature (Cold vs Hot) and their interaction. Sex and age were included in preliminary models and excluded from the final models when they were not significant. For all univariate models, clutch identity was included as a random factor. Mitochondrial related variables where log-transformed when necessary and all were mean-centered and standardized by dividing them by two times the standard deviation [@gelman2008scaling]. These variables were considered to follow a normal distribution.

Learning was modeled as a function of trial, cort, temperature, and their three-way interaction. We included in this model the trial within each level of lizard identity as a random slope. The error structure was modelled using a negative binomial distribution with a logit link function [negbinomial(link = "log")]. Otherwise, the procedure was as in the other models.

We used the posterior distributions of parameters from these models to test for differences between treatments. Learning slopes were obtained using the ‘trial’ estimates and its interaction with hormone and temperature treatments. Slope estimates less than 0 provide evidence of learning. _pMCMC_ was used to test the hypothesis that posterior distributions of slopes, and slope contrasts, were different from zero. We considered an effect statistically significant when _pMCMC_ <0.05.

To explore direct and indirect links between mitochondrial function and learning, we fitted a multivariate structural equation model (SEM) using _brms_. The model was structured based on theoretical expectations shown in @fig-sem. Any missing data was imputed during model fitting using data augmentation [@noble2021planned], but this was largely restricted to DNA damage and lipid peroxidation. To obtain a measure of learning performance we extracted the posterior distribution of individual learning slopes (i.e., changes in error over time) for each lizard. This was achieved by including a random trial slope for each lizard. Learning slopes, as well as the rest of variables, were standardized as before (see above). We assumed a gaussian error distribution for all the variables in this model. Factors found as non-significant in the univariate models were excluded. Clutch identity remained a random factor in our SEM model. Direct, indirect, and total effects were derived from posterior estimates [@kline2005principles] (see table S4).

@@

We started with 80 lizards - 20 per treatment - coming from `r total_clutches` clutches. However, due to natural mortality, only 79 lizards were included in the learning analyses, and 78 in the mitochondrial analyses. We provide final sample sizes in all our figures.

@@

Our SEM showed that mitochondrial physiology was not related to learning abilities (see @fig-sem and table S4). While we found that ROS production increased with metabolic capacity (β = `r format_dec(mean(coeff_potential_ros), 3)`, 95% CI = [`r format_dec(quantile(coeff_potential_ros, 0.025), 3)`, `r format_dec(quantile(coeff_potential_ros, 0.975), 3)`], _pMCMC_ `r format_p(pmcmc(coeff_potential_ros), 3, equal = TRUE)`), ROS was not related to  oxidative damage (see @fig-sem and table S4).

```{r, fig-sem}
#| label: fig-sem
#| fig-cap: "Structural equation model testing hypothesized direct and indirect effects of physiology on learning. Arrows indicate the directionality of the estimate, and the values show the mean predicted coefficient and its 95% confidence interval."
#| fig-name: "fig-sem"
# 
# A) Getting all the direct coefficients
density_learn <- paste0(format_dec(mean(coeff_mitodensity_learn), 3),
                  " [", format_dec(quantile(coeff_mitodensity_learn, 0.05), 3),
                  ", ", format_dec(quantile(coeff_mitodensity_learn, 0.95), 3), "]")
potential_learn <- paste0(format_dec(mean(coeff_potential_learn), 3),
                  " [", format_dec(quantile(coeff_potential_learn, 0.05), 3),
                  ", ", format_dec(quantile(coeff_potential_learn, 0.95), 3), "]")
dna_learn <- paste0(format_dec(mean(coeff_dna_learn), 3),
                  " [", format_dec(quantile(coeff_dna_learn, 0.05), 3),
                  ", ", format_dec(quantile(coeff_dna_learn, 0.95), 3), "]")
perox_learn <- paste0(format_dec(mean(coeff_perox_learn), 3),
                  " [", format_dec(quantile(coeff_perox_learn, 0.05), 3),
                  ", ", format_dec(quantile(coeff_perox_learn, 0.95), 3), "]")
#
ros_dna <- paste0(format_dec(mean(coeff_ros_dna), 3),
                  " [", format_dec(quantile(coeff_ros_dna, 0.05), 3),
                  ", ", format_dec(quantile(coeff_ros_dna, 0.95), 3), "]")
#
ros_perox <- paste0(format_dec(mean(coeff_ros_perox), 3),
                  " [", format_dec(quantile(coeff_ros_perox, 0.05), 3),
                  ", ", format_dec(quantile(coeff_ros_perox, 0.95), 3), "]")
#
density_ros <- paste0(format_dec(mean(coeff_mitodensity_ros), 3),
                  " [", format_dec(quantile(coeff_mitodensity_ros, 0.05), 3),
                  ", ", format_dec(quantile(coeff_mitodensity_ros, 0.95), 3), "]")
potential_ros <- paste0(format_dec(mean(coeff_potential_ros), 3),
                  " [", format_dec(quantile(coeff_potential_ros, 0.05), 3),
                  ", ", format_dec(quantile(coeff_potential_ros, 0.95), 3), "]")
#
imgSEM <- readPNG(here("Others", "SEM.png"))
plot_SEM <- rasterGrob(imgSEM, interpolate = TRUE)
#
fig_SEM <- ggdraw(plot_SEM) +
  annotate("text", x = 0.433, y = 0.19, label = density_learn, hjust = 1, vjust = 1, size = 3.5, family = "Times") +
  annotate("text", x = 0.433, y = 0.85, label = potential_learn, hjust = 1, vjust = 1, size = 3.5, family = "Times") +
  annotate("text", x = 0.877, y = 0.675, label = dna_learn, hjust = 1, vjust = 1, size = 3.5, family = "Times") +
  annotate("text", x = 0.877, y = 0.387, label = perox_learn, hjust = 1, vjust = 1, size = 3.5, family = "Times") +
  annotate("text", x = 0.547, y = 0.745, label = ros_dna, hjust = 1, vjust = 1, size = 3.5, family = "Times") +
  annotate("text", x = 0.547, y = 0.295, label = ros_perox, hjust = 1, vjust = 1, size = 3.5, family = "Times") +
  annotate("text", x = 0.325, y = 0.284, label = density_ros, hjust = 1, vjust = 1, size = 3.5, family = "Times") +
  annotate("text", x = 0.324, y = 0.757, label = potential_ros, hjust = 1, vjust = 1, size = 3.5, family = "Times")
ggsave(here("./output/figures/text/SEM.png"), plot = fig_SEM, width = 21, height = 10, units = "cm", dpi = 600, bg = "white")
knitr::include_graphics("./output/figures/text/SEM.png")
```

## Discussion
Lizards were able to learn a complex spatial task with marked improvements over time in their ability to choose the correct location. However, learning was not influenced by prenatal conditions or linked to mitochondrial function in the medial cortex, counter to what we predicted. Despite ROS increasing with greater metabolic capacity as expected, this did not result in a corresponding increase in oxidative damage in the medial cortex in _L. delicata_. We discuss our results in relation to existing work exploring the underlying mechanisms of learning and how early environments impact it below.
